\section*{Acknowledgements}
\addcontentsline{toc}{section}{Acknowledgements}

There has been a tremendous number of people without whom I wouldn't have made
it through my Master's and I will probably miss someone trying to list them.

I should definitely begin with my main colleague and friend -- Max Kochurov.
Most of my projects either have been done in collaboration with Max or were
otherwise inspired by his work. Throughout the master's, Max would be the first
person to discuss new ideas with. Max also has shared with me much of the
frustration related to research and to things that research makes us miss.
Thank you, Max.

Next, I must mention Dmitry Ulyanov and Thibaut Le Gouic who had been
supervising me at different times. Thibaut, you obviously were too good for me
back when I first reached out to you. I just didn't know enough -- but hey,
I've learned a lot of geometry since, and perhaps one day I'll comprehend your
works as well.  You did inspire me, however. Dmitry... we met right after I
finally broke down, and I cannot begin to describe how important it was to hear
that someone like you actually believed in me. Even before I reported
you to Skoltech as my new supervisor (which can't remember whether we ever
discussed), you had done for me as a supervisor more than anyone else:
pushing me towards research, helping with opportunities... I'm aware of the
time investment and personal effort that you've put into me, and I know that so
far I've been failing you. I'll try better. Thank you.

There have been other fellow students whom I owe helpful discussions and
inspiration. Teaching Assistant whose homework I expressively hated the most,
Alexey Boyko -- discussions with you were some of the most insightful I had!
I've been thinking about it, and perhaps you're right -- Theoretical Physics is
the right direction to converge in. I also hope we both learn to communicate
properly.
Anna Araslanova -- listen, \emph{you're great}, really. Literally at
everything.
Nick Stulov helped the SLT group not to fall apart.
Alexey Ustimenko -- you're the mean guy, but smart, and inspiring.
Alexander Lyzhov -- you're the good guy, and smart, and inspiring.
Alexey Notchenko, we haven't talked much, but when we did, that was
helpful -- at least to me. Oleg Kachan repeatedly reinforced interest in
Riemannian optimization beyond hyperbolic neural networks and helped with
opportunities to talk about my work and get feedback (even though I wasn't
particularly good at talking).
Pavel Kaloshin was one of the collaborators in our first course project and it
was because of him that I got interested in vocal.
Maria Kolos, you taught me some really important lessons that I'm still
struggling to comprehend.
Rasoul Karimov did a \emph{lot} of work on geoopt and running the experiments.
Cyrill Mazur at some point suggested me some important thoughts.
Max Brazhnikov, I think it's just culture code.

I should thank Skoltech and HSE Professors -- at least for the tolerance they
demonstrated replying to my comments and emails. Yury Maximov gave us the best
optimization course ever: I never before of a course where students would
attend all-day-long lecture followed by a seminar only to proceed the
discussion after class in the pub. I thank Bernstein for comments and
discussions regarding Riemannian optimization. Ivan Oseledets gave us advise on
our projects throughout the two years, and also provided helpful feedback on
this thesis during pre-defenses. \emph{Professor} Max Panov coordinated
us, students, throughout all of the Skoltech journey, I'm afraid to imagine how
many unread messages you have in Telegram at the end of the day. As much credit
goes to Alexey Naumov.  I thank Grigory Kabatyanski for our discussion on the
interview day -- I couldn't believe a researcher of your level would actually
listen to my presentation and even ask relevant questions. HSE's program
managers, Kristina Koshurova and Ilona Yakovleva, -- you just kept saving me
after all my mistakes and failed deadlines. Prof. Burnaev, it didn't work out,
but thanks for actually replying.

I should separately mention the Deep Bayes group whom I owe their wonderful
seminars, discussions in a pub, and the Neurobayesian methods course. Dmitry Vetrov
and Kyrill Neklyudov provided Max and me with important feedback in our work on
hyperbolic convolutions. Dmitry Kropotov provided with rather conceptual
discussions of Riemannian optimization. I thank Dmitry Molchanov for the discussion
during an NBM seminar. I thank my two other NBM examiners, Oleg Struminsky and Art
Sobolev for their patience, for the discussions, and for not letting me get away
\emph{too} easily. Rest assured, I'll master your Bayesian discipline, I
already have some ideas as to where I might need it. I thank Alexander Grishin
for the NBM course and the reading group at Yandex.

Gary and Octavian worked a lot with Max and helped us in hyperbolic convolutions
research.

The in3D team -- I remember when it was a full team of fighters! Nikolai,
Vadim, Vsevolod, Ilya, Bulat, and, of course, Dmitry -- you're just great.
The impact you had on what I choose to research would be... obvious.

Mentioning in3D, I should also say that the roots of my interest towards this field
belong in Russian3DScanner.

I must mention my friends, without which making it to the end wouldn't have
been possible. Eugene, probably my oldest friend -- just... thank you for being
there. And also for reinforcing the interest in engineering with cool links and chats.
All of Ziferblat -- I owe you! Diverse discussions with Abdalaziz Rashid
and Arseny. Casual category theory, functional programming, formal
theorem proving, structure from motion, and experimental dynamics of paper
airplanes -- Alexander Lebedev was a huge source of inspiration. I thank Liza
and Fed Lidsky for the discussions, for bicycle nights, for Metallica, and the
Modenirst Cuisine. I thank Jonathan for the funniest French lessons, for Muay
Thai, and just for being a great guy. Dunya, I'm sorry for not reaching out,
that just wasn't a good time. Liliya Kondrasheva, I think we found the same thing
in Ziferblat. Liza Chesnokova, Katya Dubrovina, Varya Ozhiganova, Sasha
Tikhonova, Zhenya, Stas... Elya Ortach, the museum raids we organized, finishing
at De Faqto. Mila, the church organ -- we should really make it. Dinara, glad
you've made it, I'll now try to make it too. Bogucharsky! The greatest thing
about you is that it's \emph{so obvious} how excited you are by the job you're doing.
Kostya Karp, Tula happened because of you. I've almost committed a crime
forgetting to mention Katya Dagkesamanskaya, whom I should thank for sharing
the passion for languages, theatre, music, and science.
There are too many more names to
mention, and I keep avoiding the ones I think of first. Anna, that ticket one
hour before the train was a life changer: I had discovered a lot both about
myself and about the world. Now that the worst is over, I am remembering the
inspiring. Above all, I'm remembering someone who has a vision and lives by it.
Alex, the Petropavlovsky forest, the ride from Danilovsky, the snow in the
Milyutinsky lane, Bolotov, Voronez, and Volga -- I don't know anyone else so
easy to take off and to turn everything around into magic. You've been my
closest friend, and that friendship was the healthiest relationship I ever had.
Olya, the reassurance you gave cannot be underestimated.
Nikita and Feodor -- well, you know.

I should thank Masha Sheyanova, Vadim Lebedev, Vika Chekalina, Danya
Alexeyevsky, and Satyarth for the amazing time we had, and for ``inverting the
sphere inside out''

I thank Valery Kharitonov for help and inspiration both throughout my
Bachelor's and Master's.

Anatoly Baskakov, Yegor Dikarev, Ilya Krishtal, Lilya Kachaturovna -- I lost
the contact lately, but your influence on me is permanent.

Finally, I should beg indulgence of my family for I've been missing.
